---
title: "Assignment 2: Coding Basics"
author: "Rachel Gordon"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. Use the seq command to get a sequence of numbers from 1 to 100 at intervals of 4
seq(1,100,4)

#Assigning sequence above the name sequence.a
sequence.a<-seq(1,100,4)

#2. Compute the mean and median of this sequence
mean(sequence.a)
median(sequence.a)

#3. Determining if the mean of sequence.a is greater than the median
mean(sequence.a)>median(sequence.a)
```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
# create a series of vectors
a <- c("Rachel", "Sara", "John", "Michael") #character vector
b <- c(87,43,90,82) #numeric vector
c <- c(TRUE,FALSE,TRUE,TRUE) #logical vector

# Combining vectors into a data frame, assigning the data frame a name
student.scores <-data.frame(a,b,c) #combining vectors into a data frame and naming data frame

#label columns of data frame
names(student.scores) <-c("Name","Score","Passed"); View(student.scores)

print(student.scores)

```

9. QUESTION: How is this data frame different from a matrix?

> Answer: This data frame is different from a matrix because in this data frame, there is data of different types in each column. In a matrix, however, all the data would be of the same data type.

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
# Create a function with an if/else statement to indicate True/False for a passing grade more than 50

passing.test.score <- function(x) {
       ifelse(x >= 50,"TRUE","FALSE")
       }

# Apply function to the test score vector
passing.test.score(b)

# Create a function with 'if' and 'else' statement to indicate True/False for a passing grade more than 50
passing.test.score2 <- function(x) {
  if(x >= 50) {
    TRUE
  } 
  else {
    FALSE
  }
}

# Apply function to the test score vector
passing.test.score2(b)

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 'ifelse' worked. This is because 'ifelse' is used to check every element of the vector, whereas using just 'if'  and 'else' will only check the first element of the vector. As shown above, using 'if' and 'else' creates a warning message that only the first element will be used and then only returns the data for the first element of the vector when the function is applied.
